Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-11-01 manuscript/climate_sensitivity_draft.Rmd
working on references

Our results reveal how tree size, microhabitat, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The tendencey for larger trees to suffer more, observed here as in forests around the world [@bennett_larger_2015], was driven primarily by their height. There was a marginal additional effect of crown exposure, with the most exposed and the most suppressed trees suffering most--consistent with observations of both greater drought sensitivity of exposed trees (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and greater sensitivity of suppressed and crowded individuals (**REFS**). There was no evidence that root water access increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [@zuleta_drought-induced_2017]), and larger rooting volume provided no advantage in the drier microenvironments. The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species. Drought-sensitive species were not linked to $LMA$ and $WD$ but were predicted--at least some of the time--by hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$, and $XP$), which is physiologically logical (**REFS**[@kannenberg_linking_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of these responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most. These findings significantly advance our knowledge as to the factors that confer vulnerability or resilience on trees during drought.

The droughts considered here were of similar severity (Fig. 1b) and fairly moderate; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019). Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature. While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), underlining the fact that these droughts did not induce extreme stress on the entire forest. It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 10-12% for all droughts combined, and 21-27% for each individual drought (Table 5). Methodologically, the moderate nature of these drougths is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed. These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*). Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent-- traits (**REFS**). Nevertheless, our results are consistent with findings from more extreme droughts.

Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height. To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (**REFS**). They also face different microenvironments, which are partially related to canopy position. Even under non-drought conditions, evaporative demand increases with tree height, which are more closely coupled to the atmosphere  (Fig. 1a-b; **REFS- Jarvis 1984?**; [@bretfeld_plant_2018]). **The vegetation reaches higher temperatures[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)**, which would occur more during a drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. Furthermore, daytime CO_2 concentrations tend to be lower in the canopy than in the understory **(Fig. 2[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)** [@koike_leaf_2001; MORE REFS], implying greater water costs of $CO_2$ uptake. **[McDowell-vertical profiles in isotopes?*]**. Taller trees are more likely to be in dominant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig. 2**e?**) [@muller-landau_testing_2006]. We show that height is a far stronger predictor than crown position (Tables 1,4,5). Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4). However, rates of change are relatively slow, and the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of (*H1.2*), implying that dominant crown positions did indeed reduce $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??).  It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer distproportionately during drought (**REFS**). The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.

>>>>

Our results reveal how tree size, microhabitat, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The tendencey for larger trees to suffer more, observed here as in forests around the world [@bennett_larger_2015], was driven primarily by their height. There was a marginal additional effect of crown exposure, with the most exposed and the most suppressed trees suffering most--consistent with observations of both greater drought sensitivity of exposed trees (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and greater sensitivity of suppressed and crowded individuals (**REFS**). There was no evidence that root water access increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [@zuleta_drought-induced_2017]), and larger rooting volume provided no advantage in the drier microenvironments. The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species. Drought-sensitive species were not linked to $LMA$ and $WD$ but were predicted--at least some of the time--by hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$, and $XP$), which is physiologically logical ([@scoffoni_leaf_2014];[@bartlett_correlations_2016]; [@medeiros_extensive_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of these responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most. These findings significantly advance our knowledge as to the factors that confer vulnerability or resilience on trees during drought.

The droughts considered here were of similar severity (Fig. 1b) and fairly moderate; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019). Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature. While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), underlining the fact that these droughts did not induce extreme stress on the entire forest. It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-27% for each individual drought (Table 5). Methodologically, the moderate nature of these droughts is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed. These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*). Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits (*Greenwood et al. 2007- doi: 10.1111/ele.12748; Anderegg et al. 2016- doi: 10.1073/pnas.1525678113*). Nevertheless, our results are consistent with findings from more extreme droughts.

Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height. To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (*Ryan et al. 2006-  DOI:10.1111/j.1365-3040.2005.01478.x; McDowell & Allen 2015: DOI: 10.1038/NCLIMATE2641; [McDowell et al. 2011](https://www.springer.com/gp/book/9789400712416)*). They also face different microenvironments, which are partially related to canopy position. Even under non-drought conditions, evaporative demand increases with tree height, which are more closely coupled to the atmosphere  (Fig. 1a-b; **REFS- Jarvis 1984?**; [@bretfeld_plant_2018]). Exposed canopy leaves reach higher temperatures (**REFS**), particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. Furthermore, daytime CO_2_ concentrations tend to be lower in the canopy than in the understory [@koike_leaf_2001], implying greater water costs of $CO_2$ uptake. Taller trees are more likely to be in dominant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig. 2d) [@muller-landau_testing_2006]. We show that height is a far stronger predictor than crown position (Tables 1,4,5). Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4). However, rates of change are relatively slow, and the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of *H1.2*. The implication is that dominant crown positions did have a marginally negative influence on $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]). It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS**). The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.
